Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-01 paper/paper.tex
Filled in some of the implementation section.

@@

>>>>

The first four techniques reside in the purely functional realm. The
following three techniques use imperative constructs to avoid allocation
overhead.

@@

% For David to fill in.
\subsection{Imperative global store and worklist}
...
\subsection{Pre-allocated global store}
...

% Ian's sections: generators, implementation and compound data
\section{Implementation}

All the above techniques can be combined into a single implementation
of the small step reduction relation and small variations of a general
fixpoint combinator. We implementated all of these ideas in an
analysis of R4RS scheme in just 3000 lines of Racket. Most lines are
spent listing all the primitives and library functions (and their
abstract and concrete implementations) of R4RS. Here we describe the
unifying vocabulary for writing such a framework, and some design
decisions related to scaling to a language with compound data.

\subsection{Vocabulary of Parameterized Optimizations}

We combine the interface of generators ~\cite{ianjohnson:cluhistory}
with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
of {\tt ev} states. We also have a special {\tt do} form that performs
behind-the-scenes state passing and iteration much like {\tt do} in
Haskell. In the presence of the different optimization
techniques, {\tt generator}, {\tt yield}, {\tt $\lambda$\%} and {\tt do}
change meaning slightly, but keep the same interface as their names
suggest.

\begin{figure}
\begin{alltt}
\(e-dispatch =\)
 (match e
    [(var\(\superscript\ell\) x)
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) ((v (delay (lookup-env \(\menv\) x))))
        (yield (co \(\msto\) k v))))]
    [(lit\(\superscript\ell\) l)
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
    [(lam\(\superscript\ell\) x e)
     (define c (compile e))
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
    [(app\(\superscript\ell\) f e)
     (define cf (compile f))
     (define ce (compile e))
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (define-values (\(\msto\)* a) (push state))
      (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
    [(ife\(\superscript\ell\) e0 e1 e2)
      (define c0 (compile e0))
      (define c1 (compile e1))
      (define c2 (compile e2))
      (\(\lambda\)\% (\(\msto\) \(\menv\) k)
        (do (\(\msto\)) [(\(\msto\)* a) (push state)]
         (yield (ev \(\msto\)* c0 \(\menv\) (ifk c1 c2 \(\menv\) a)))))])

  (define (step state)
    (match state
      [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
      [(co \(\msto\) k v)
       (match k
         ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
           (yield (ans \(\msto\) v))))]
         [(ar\(\superscript\ell\) e \(\menv\))
          (generator (do (\(\msto\)) ()
            (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
         [(fn\(\superscript\ell\) f)
          (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
           (yield (ap \(\msto\) f v k))))]
         [(fi\(\superscript\ell\) c a \(\menv\))
          (generator
           (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                    [v <- (force \(\msto\) v)])
             (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
      [(ap \(\msto\) fun a k)
       (match fun
         [(clos l x e \(\menv\))
          (generator
           (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
             (yield (ev \(\msto\)* e \(\menv\)* k))))]
         [(? op? o)
          (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                   [v <- (force \(\msto\) v)]
                   [v (\(\Delta\) o (list v))])
            (yield (co \(\msto\) k v)))
         [_ (generator (do (\(\msto\)) () (continue)))]))]))
\end{alltt}
\caption{Parameterized Abstract$^2$ machine for ISWIM}
\label{fig:paam}
\end{figure}

\begin{figure}
\begin{alltt}
(define (compile e) \(e-dispatch\))
(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\machstep\) (\(\lambda\) (\(\msto\) \(\menv\) k) body \ldots)
(yield (ev \(\msto\) c . rest)) \(\longmapsto\) (c \(\msto\) . rest)
(yield e) \(\longmapsto\) (yield-meaning e)
\end{alltt}
\caption{Meaning of forms if compiled}
\label{fig:cfm}
\end{figure}

\begin{figure}
\begin{alltt}
(define (compile e) e)
(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\longmapsto\) (let () body ...)
(yield e) \(\longmapsto\) (yield-meaning e)
\end{alltt}
\caption{Meaning of forms if not compiled}
\label{fig:ncfm}
\end{figure}

Why generators? Relations in general cannot be written as computable
functions, but in our case we have a decidable, finitely
non-deterministic relation. There are several implementation
strategies for relations, some more effective than others. For
instance, a silly implementation strategy would be defining a binary
function that decides membership in the relation. We want the
definition to drive the interpretor, so we might naturally decide to
write a function that takes a state and returns a set of all the next
states. As we described in the imperative workset section, this
intermediate allocation is unnecessary and wasteful. What we really
want is a way of pulling on the reduction relation for more next
states as the fixpoint is ready for them.

For intermediate sets, {\tt yield} is {\tt return} in the set
monad. For the imperative worklist, {\tt yield} performs mutation
behind the scenes. Indeed, we could use real generators! This is the
only case where {\tt generator} constructs a nullary generator; in the
other cases it evaluates to its body. Racket's generators are not
tuned for performance, so we actually lose efficiency in this case,
but the unifying vocabulary is instructive nonetheless.

The {\tt $\lambda$\%} form constructs a real closure in the
compilation case and evaluates to its body otherwise.

The {\tt do} form has the special job of maintaining the current store
and accumulated states (if indeed those are being passed around). It
uses Racket's syntax parameters ~\cite{ianjohnson:eli/stxparam} to
track the most recent bindings of these accumulators so that {\tt
  yield-meaning} can refer to them (say {\tt $\sigma$p} and {\tt
  statep}). In the case of state passing and set-accumulated states, its
definition is simply

\begin{center}
\begin{alltt}
(yield-meaning e) \(\longmapsto\) (values \(\sigma\)p (set-add statep e))
\end{alltt}
\end{center}

The linguistic elements described above were all reusable to implement
the various languages' abstract interpreters. One such language was
R4RS Scheme, which has compound data (e.g. vectors and conses).

\subsection{Abstracting Compound Data}

We discovered that the choice of abstract domain for literal data
(such as the large quoted axiom database in the Boyer benchmark)
drastically affected the performance of our analysis on the benchmark
suite. It is worth discussing the ``obvious'' and na\"ive abstraction
that is natural to a reader of
~\citep{dvanhorn:VanHorn2011Abstracting}, and other easy-to-implement
yet effective approaches.

TODO: {\tt (cons A-addr D-addr)}, {\tt qpair}, {\tt opaque-atom}.
